---
title: "Assignment"
format: html
editor: visual
---

```{r setup, warning=FALSE, message=FALSE}
setwd("C:/Users/vezuc/OneDrive - Florida International University/PhD/Classes/Bayesian")

knitr::opts_chunk$set(warning = FALSE, message = FALSE)

suppressWarnings(suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)     
  library(stringr)
  library(lubridate)
  library(purrr)
  library(brms)
  library(cmdstanr)
  library(posterior)
  library(furrr)
  library(future)
  }))

project_dir <- "C:/Users/vezuc/OneDrive - Florida International University/PhD/Classes/Bayesian"
srt_dir <- file.path(project_dir, "SRT files")

setwd(project_dir)

```

## Read data

```{r}
meta <- readxl::read_excel("srt_metadata_with_tide_moon.xlsx", sheet = "srt_metadata") %>%
  mutate(
    file_date = as.Date(file_date),
    year = factor(lubridate::year(file_date)),
    month_num = lubridate::month(file_date),
    month = factor(month.abb[month_num], levels = month.abb, ordered = TRUE),
    day = lubridate::day(file_date),
    tide_state = factor(tide_state),
    moon_phase = factor(moon_phase)
  )

srt_files <- list.files(
  path = srt_dir,
  pattern = "\\.SRT$",
  full.names = TRUE,
  recursive = TRUE
)
```

## Function to parse one SRT file

```{r}
parse_srt_gps_fast <- function(file_path) {
  txt <- readr::read_file(file_path)
  
  blocks <- strsplit(txt, "\\r?\\n\\r?\\n")[[1]]
  blocks <- blocks[stringr::str_detect(blocks, "-->")]
  
  if (length(blocks) == 0) return(NULL)
  
  time_match <- stringr::str_match(
    blocks,
    "(\\d{2}):(\\d{2}):(\\d{2})[,\\.](\\d{3})\\s+-->"
  )
  
  elapsed_sec <- as.numeric(time_match[,2]) * 3600 +
    as.numeric(time_match[,3]) * 60 +
    as.numeric(time_match[,4]) +
    as.numeric(time_match[,5]) / 1000
  
  lat1 <- stringr::str_match(blocks, "[Ll]atitude[: ]+(-?\\d+\\.\\d+)")
  lon1 <- stringr::str_match(blocks, "[Ll]ongitude[: ]+(-?\\d+\\.\\d+)")
  
  gps2 <- stringr::str_match(
    blocks,
    "GPS\\s*\\(?\\s*(-?\\d+\\.\\d+)\\s*,\\s*(-?\\d+\\.\\d+)"
  )
  
  latitude <- ifelse(!is.na(lat1[,2]), as.numeric(lat1[,2]), as.numeric(gps2[,3]))
  longitude <- ifelse(!is.na(lon1[,2]), as.numeric(lon1[,2]), as.numeric(gps2[,2]))
  
  dt_match <- stringr::str_match(
    blocks,
    "(20\\d{2}-\\d{2}-\\d{2})\\s+(\\d{2}:\\d{2}:\\d{2})"
  )
  
  datetime_srt <- suppressWarnings(
    lubridate::ymd_hms(paste(dt_match[,2], dt_match[,3]), quiet = TRUE)
  )
  
  out <- tibble::tibble(
    file_name = basename(file_path),
    elapsed_sec = elapsed_sec,
    latitude = latitude,
    longitude = longitude,
    datetime_srt = datetime_srt
  ) %>%
    dplyr::filter(!is.na(latitude), !is.na(longitude), !is.na(elapsed_sec))
  
  if (nrow(out) == 0) return(NULL)
  
  out
}

test1 <- parse_srt_gps_fast(srt_files[1])
test1
dim(test1)
head(test1)

## Extract GPS from all SRT files
# set up parallel processing
plan(multisession, workers = 6)

# folder where one CSV per SRT file will be saved
dir.create("gps_srt_csvs", showWarnings = FALSE)

# optional: sort file list for consistency
srt_files <- sort(srt_files)

# function to process one SRT file and save one CSV per file
process_one_srt <- function(f) {
  
  # create a unique output name based on the full file path
  file_id <- normalizePath(f, winslash = "/") %>%
    stringr::str_remove("\\.[Ss][Rr][Tt]$") %>%
    gsub("[^A-Za-z0-9_]+", "_", .)
  
  out_file <- file.path("gps_srt_csvs", paste0(file_id, ".csv"))
  
  # skip if this file has already been processed
  if (file.exists(out_file)) {
    message("Skipping already processed file: ", basename(f))
    return(NULL)
  }
  
  message("Processing: ", basename(f))
  
  # parse SRT
  out <- tryCatch(
    parse_srt_gps_fast(f),
    error = function(e) {
      message("Failed: ", basename(f), " | Error: ", e$message)
      return(NULL)
    }
  )
  
  # save result
  if (!is.null(out) && nrow(out) > 0) {
    write_csv(out, out_file)
  } else {
    # save empty file so failed/empty files are not retried endlessly
    write_csv(tibble(), out_file)
  }
  
  return(NULL)
}

# process all SRT files
future_walk(srt_files, process_one_srt)

# combine all per-file CSVs into one dataset
csv_files <- list.files(
  "gps_srt_csvs",
  pattern = "\\.csv$",
  full.names = TRUE
)

gps_raw <- csv_files %>%
  map_dfr(function(x) {
    df <- read_csv(x, show_col_types = FALSE)
    if (nrow(df) == 0) return(NULL)
    df
  })

# optional: save combined dataset
write_csv(gps_raw, "gps_raw_combined.csv")

# return to sequential processing
plan(sequential)
```

```{r}
##Sanity check
head(srt_files)

dim(gps_raw)
head(gps_raw)
summary(gps_raw$latitude)
summary(gps_raw$longitude) 

gps_raw %>%
  count(file_name) %>%
  arrange(desc(n)) %>%
  head()

summary(is.na(gps_raw$datetime_srt))
```

## Keep one GPS point every 2 seconds

```{r}
gps_2s <- gps_raw %>%
  mutate(time_bin_2 = floor(elapsed_sec / 2) * 2) %>%
  group_by(file_name, time_bin_2) %>%
  slice(1) %>%
  ungroup()

##sanity check
dim(gps_2s)
head(gps_2s)
summary(gps_2s$latitude)
summary(gps_2s$longitude)

##save data
write_csv(gps_2s, "gps_2s.csv")
```

## Raw path map (2 sec intervals)

```{r}
ggplot(gps_2s, aes(x = longitude, y = latitude, group = file_name)) +
  geom_path(alpha = 0.25, linewidth = 0.2, color = "steelblue4") +
  coord_equal() +
  theme_bw() +
  labs(
    x = "Longitude",
    y = "Latitude",
    title = "Tiger shark paths (2-second intervals)"
  )
```

## Hotspot map (2 second interval)

```{r}
ggplot(gps_2s, aes(x = longitude, y = latitude)) +
  stat_density_2d_filled(contour_var = "density") +
  coord_equal() +
  theme_bw() +
  labs(
    x = "Longitude",
    y = "Latitude",
    title = "Tiger shark hotspot areas (2-second intervals)"
  )
```

## Step distance check

```{r}
##Step distance check to see if 2-second data was capturing real movement, or just noise

gps_2s %>%
  group_by(file_name) %>%
  summarise(
    step_dist = mean(sqrt(diff(longitude)^2 + diff(latitude)^2), na.rm = TRUE)
  ) %>%
  summary()
```

## GPS figure for every 2 second intervals

```{r}
library(sf)
library(dplyr)
library(purrr)
library(ggplot2)
library(maptiles)
library(tidyterra)

##Make SF object
gps_2s_sf <- gps_2s %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)

```

## Line object for shark paths

```{r}
line_df <- gps_2s %>%
  arrange(file_name, elapsed_sec) %>%
  group_by(file_name) %>%
  filter(n() >= 2) %>%
  group_split()

line_list <- map(line_df, ~ st_linestring(as.matrix(.x[, c("longitude", "latitude")])))

shark_lines <- st_sf(
  file_name = map_chr(line_df, ~ .x$file_name[1]),
  geometry = st_sfc(line_list, crs = 4326)
)
```

## Build a bounding box around the shark tracks

```{r}
bbox_sf <- st_as_sfc(st_bbox(gps_2s_sf)) %>%
  st_buffer(0.0008)
```

## Download satellite basemap

```{r}
sat_map <- get_tiles(
  x = bbox_sf,
  provider = "Esri.WorldImagery",
  crop = TRUE,
  zoom = 17
)
```

## Static hotspot map on satellite imagery

```{r}
ggplot() +
  geom_spatraster_rgb(data = sat_map) +
  stat_density_2d_filled(
    data = gps_2s,
    aes(x = longitude, y = latitude),
    contour_var = "density",
    alpha = 0.42
  ) +
  coord_sf(
    xlim = st_bbox(bbox_sf)[c("xmin", "xmax")],
    ylim = st_bbox(bbox_sf)[c("ymin", "ymax")],
    expand = FALSE
  ) +
  theme_void() +
  labs(
    title = "Tiger shark hotspot areas",
    fill = "Relative use",
    caption = "Basemap: Esri World Imagery"
  )
```

## Flag/ remove impossible GPS points

```{r}
library(sf)
library(dplyr)
library(ggplot2)
library(maptiles)
library(tidyterra)
```

```{r}
gps_2s_sf <- gps_2s %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)
```

### Import water polygon

```{r}
water_poly <- st_read("Sueste_water_polygon.kml")
```

### Use water polygon to flag GSP points on land

```{r}
gps_2s_sf <- gps_2s %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)

gps_2s_sf <- gps_2s_sf %>%
  mutate(in_water = lengths(st_within(geometry, water_poly)) > 0)
```

### Check flagging manually

```{r}
ggplot() +
  geom_sf(data = water_poly, fill = "lightblue", alpha = 0.2, color = "blue") +
  geom_sf(data = gps_2s_sf, aes(color = in_water), size = 0.4, alpha = 0.5) +
  theme_bw()
```

## Data cleaning: Keep only water point

```{r}
gps_2s_clean <- gps_2s_sf %>%
  filter(in_water) %>%
  st_drop_geometry()

write_csv(gps_2s_clean, "gps_2s_clean.csv")
```

```{r}
##hotspot map
ggplot(gps_2s_clean, aes(x = longitude, y = latitude)) +
  stat_density_2d_filled(contour_var = "density") +
  coord_equal() +
  theme_bw() +
  labs(
    x = "Longitude",
    y = "Latitude",
    title = "Tiger shark hotspot areas (water-only points)"
  )

##satellite hotspot map
ggplot() +
  geom_spatraster_rgb(data = sat_map) +
  stat_density_2d_filled(
    data = gps_2s_clean,
    aes(x = longitude, y = latitude),
    contour_var = "density",
    alpha = 0.30,
    bins = 6
  ) +
  geom_sf(
    data = water_poly,
    fill = NA,
    color = "cyan",
    linewidth = 0.6
  ) +
  coord_sf(
    xlim = st_bbox(bbox_sf)[c("xmin", "xmax")],
    ylim = st_bbox(bbox_sf)[c("ymin", "ymax")],
    expand = FALSE
  ) +
  theme_void() +
  labs(
    title = "Tiger shark core-use areas",
    fill = "Relative use",
    caption = "Basemap: Esri World Imagery"
  )

```

### Save flagged data

```{r}
gps_2s_flagged <- gps_2s_sf %>%
  mutate(location_flag = if_else(in_water, "water", "land")) %>%
  st_drop_geometry()

write_csv(gps_2s_flagged, "gps_2s_flagged.csv")
```

## Stan model with one track only

```{r}
##prepare
library(dplyr)
library(sf)
library(cmdstanr)
library(posterior)
library(ggplot2)

one_file <- gps_2s_clean %>%
  filter(file_name == unique(gps_2s_clean$file_name)[1]) %>%
  arrange(elapsed_sec) %>%
  slice(1:300)

## convert to sf and project to meters
one_file_sf <- one_file %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE) %>%
  st_transform(32725)

## Extract x/y in meters
coords_m <- st_coordinates(one_file_sf)

one_file <- one_file %>%
  mutate(
    x_m = coords_m[, 1],
    y_m = coords_m[, 2]
  )

## center the meter coordinates
one_file <- one_file %>%
  mutate(
    x_obs = x_m - mean(x_m),
    y_obs = y_m - mean(y_m)
  )

## build Stan data
track_dat <- list(
  N = nrow(one_file),
  x_obs = one_file$x_obs,
  y_obs = one_file$y_obs
)

## CRW Stan model in meters
writeLines("
data {
  int<lower=3> N;
  vector[N] x_obs;
  vector[N] y_obs;
}
parameters {
  vector[N] x_true;
  vector[N] y_true;
  real<lower=0, upper=1> phi;
  real<lower=0> sigma_gps;
  real<lower=0> sigma_move;
}
transformed parameters {
  vector[N-1] dx_true;
  vector[N-1] dy_true;

  for (t in 2:N) {
    dx_true[t-1] = x_true[t] - x_true[t-1];
    dy_true[t-1] = y_true[t] - y_true[t-1];
  }
}
model {
  phi ~ beta(2, 2);

  sigma_gps ~ exponential(1);
  sigma_move ~ exponential(1);

  x_true[1] ~ normal(x_obs[1], 1);
  y_true[1] ~ normal(y_obs[1], 1);

  x_true[2] ~ normal(x_obs[2], 1);
  y_true[2] ~ normal(y_obs[2], 1);

  for (t in 2:(N-1)) {
    dx_true[t] ~ normal(phi * dx_true[t-1], sigma_move);
    dy_true[t] ~ normal(phi * dy_true[t-1], sigma_move);
  }

  x_obs ~ normal(x_true, sigma_gps);
  y_obs ~ normal(y_true, sigma_gps);
}
generated quantities {
  vector[N] x_rep;
  vector[N] y_rep;

  for (t in 1:N) {
    x_rep[t] = normal_rng(x_true[t], sigma_gps);
    y_rep[t] = normal_rng(y_true[t], sigma_gps);
  }
}
", "shark_crw_meters.stan")


## Fit the model
mod_crw <- cmdstan_model("shark_crw_meters.stan")

fit_crw <- mod_crw$sample(
  data = track_dat,
  chains = 4,
  iter_warmup = 1000,
  iter_sampling = 1000,
  refresh = 0
)

## check parameters estimates
fit_crw$summary(variables = c("phi", "sigma_gps", "sigma_move"))

## extract corrected path in meters
x_post <- fit_crw$draws("x_true", format = "matrix")
y_post <- fit_crw$draws("y_true", format = "matrix")

x_mean <- colMeans(x_post) + mean(one_file$x_m)
y_mean <- colMeans(y_post) + mean(one_file$y_m)

track_fit <- one_file %>%
  mutate(
    x_true_m = x_mean,
    y_true_m = y_mean
  )

## convert corrected path back to lon/lat
x_post <- fit_crw$draws("x_true", format = "matrix")
y_post <- fit_crw$draws("y_true", format = "matrix")

x_mean <- colMeans(x_post) + mean(one_file$x_m)
y_mean <- colMeans(y_post) + mean(one_file$y_m)

track_fit <- one_file %>%
  mutate(
    x_true_m = x_mean,
    y_true_m = y_mean
  )


##plot in meters
ggplot() +
  geom_path(
    data = track_fit,
    aes(x = x_m, y = y_m),
    color = "grey60",
    linewidth = 0.5,
    alpha = 0.8
  ) +
  geom_path(
    data = track_fit,
    aes(x = x_true_m, y = y_true_m),
    color = "red",
    linewidth = 0.8
  ) +
  coord_equal() +
  theme_bw() +
  labs(
    x = "X (meters)",
    y = "Y (meters)",
    title = "Observed and Bayesian-corrected shark path (CRW, meters)"
  )


## plot observed vs corrected in meters
ggplot() +
  geom_path(
    data = track_fit,
    aes(x = x_m, y = y_m),
    color = "grey60",
    linewidth = 0.5,
    alpha = 0.8
  ) +
  geom_path(
    data = track_fit,
    aes(x = x_true_m, y = y_true_m),
    color = "red",
    linewidth = 0.8
  ) +
  coord_equal() +
  theme_bw() +
  labs(
    x = "X (meters)",
    y = "Y (meters)",
    title = "Observed and Bayesian-corrected shark path (CRW, meters)"
  )

```

## Quantify correction magnitude

```{r}
track_fit <- track_fit %>%
  mutate(
    correction_dist_m = sqrt((x_m - x_true_m)^2 + (y_m - y_true_m)^2)
  )

summary(track_fit$correction_dist_m)
```

## Batch processing using meters-based CRW Stan model

```{r}
library(dplyr)
library(sf)
library(posterior)
library(purrr)
library(readr)
library(tidyr)

fit_one_crw_track <- function(df_track, mod_crw) {
  
  df_track <- df_track %>%
    arrange(elapsed_sec)
  
  if (nrow(df_track) < 3) return(NULL)
  
  track_sf <- df_track %>%
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE) %>%
    st_transform(32725)
  
  coords_m <- st_coordinates(track_sf)
  
  df_track <- df_track %>%
    mutate(
      x_m = coords_m[, 1],
      y_m = coords_m[, 2],
      x_center = x_m - mean(x_m),
      y_center = y_m - mean(y_m)
    )
  
  track_dat <- list(
    N = nrow(df_track),
    x_obs = df_track$x_center,
    y_obs = df_track$y_center
  )
  
  fit <- mod_crw$sample(
    data = track_dat,
    chains = 4,
    iter_warmup = 1000,
    iter_sampling = 1000,
    refresh = 0
  )
  
  x_post <- fit$draws("x_true", format = "matrix")
  y_post <- fit$draws("y_true", format = "matrix")
  
  x_mean <- colMeans(x_post) + mean(df_track$x_m)
  y_mean <- colMeans(y_post) + mean(df_track$y_m)
  
  corrected_sf <- st_as_sf(
    tibble(x_true_m = x_mean, y_true_m = y_mean),
    coords = c("x_true_m", "y_true_m"),
    crs = 32725,
    remove = FALSE
  ) %>%
    st_transform(4326)
  
  coords_ll <- st_coordinates(corrected_sf)
  
  param_sum <- fit$summary(variables = c("phi", "sigma_gps", "sigma_move")) %>%
    select(variable, mean) %>%
    pivot_wider(names_from = variable, values_from = mean)
  
  out <- df_track %>%
    mutate(
      x_true_m = x_mean,
      y_true_m = y_mean,
      longitude_true = coords_ll[, 1],
      latitude_true = coords_ll[, 2],
      correction_dist_m = sqrt((x_m - x_true_m)^2 + (y_m - y_true_m)^2),
      phi = param_sum$phi,
      sigma_gps = param_sum$sigma_gps,
      sigma_move = param_sum$sigma_move
    )
  
  out
}

dir.create("crw_outputs", showWarnings = FALSE)
dir.create("crw_outputs/track_files", showWarnings = FALSE)
dir.create("crw_outputs/batch_files", showWarnings = FALSE)

track_list <- gps_2s_clean %>%
  group_by(file_name) %>%
  group_split()

batch_size <- 10
track_indices <- seq_along(track_list)
batch_ids <- split(track_indices, ceiling(seq_along(track_indices) / batch_size))

all_batch_files <- character()

for (b in 1:1) {
  
  cat("Starting batch", b, "of", length(batch_ids), "\n")
  
  this_batch_idx <- batch_ids[[b]]
  this_batch_tracks <- track_list[this_batch_idx]
  batch_results <- vector("list", length(this_batch_tracks))
  
  for (i in seq_along(this_batch_tracks)) {
    
    this_track <- this_batch_tracks[[i]]
    this_name <- this_track$file_name[1]
    safe_name <- make.names(this_name)
    track_outfile <- file.path("crw_outputs/track_files", paste0(safe_name, "_crw.csv"))
    
    if (file.exists(track_outfile)) {
      cat("  Skipping already saved track:", this_name, "\n")
      batch_results[[i]] <- read_csv(track_outfile, show_col_types = FALSE)
      next
    }
    
    cat("  Fitting track", i, "in batch", b, ":", this_name, "\n")
    
    fit_result <- tryCatch(
      fit_one_crw_track(this_track, mod_crw),
      error = function(e) {
        message("    Failed on track: ", this_name)
        message("    Error: ", e$message)
        return(NULL)
      }
    )
    
    batch_results[[i]] <- fit_result
    
    if (!is.null(fit_result)) {
      write_csv(fit_result, track_outfile)
    }
  }
  
  batch_df <- bind_rows(batch_results)
  
  batch_file <- file.path(
    "crw_outputs/batch_files",
    paste0("crw_batch_", sprintf("%03d", b), ".csv")
  )
  
  write_csv(batch_df, batch_file)
  all_batch_files <- c(all_batch_files, batch_file)
  
  cat("Finished batch", b, "- saved to:", batch_file, "\n")
}

batch_files <- list.files(
  "crw_outputs/batch_files",
  pattern = "^crw_batch_.*\\.csv$",
  full.names = TRUE
)

gps_crw_all <- batch_files %>%
  map_dfr(read_csv, show_col_types = FALSE)

write_csv(gps_crw_all, "crw_outputs/gps_2s_clean_crw_all_tracks.csv")

gps_crw_all <- gps_crw_all %>%
  mutate(
    correction_dist_m = sqrt((x_true_m - x_obs)^2 + (y_true_m - y_obs)^2)
  )

track_summary <- gps_crw_all %>%
  group_by(file_name) %>%
  summarise(
    n = n(),
    mean_correction = mean(correction_dist_m, na.rm = TRUE),
    median_correction = median(correction_dist_m, na.rm = TRUE),
    max_correction = max(correction_dist_m, na.rm = TRUE),
    .groups = "drop"
  )

write_csv(track_summary, "crw_outputs/crw_track_summary.csv")

gps_crw_all %>%
  summarise(
    mean_correction = mean(correction_dist_m, na.rm = TRUE),
    median_correction = median(correction_dist_m, na.rm = TRUE),
    max_correction = max(correction_dist_m, na.rm = TRUE)
  )

write_csv(gps_crw_all, "crw_outputs/gps_2s_clean_crw_batch1_test.csv")
```

### Run remaining batches

```{r}
track_list <- gps_2s_clean %>%
  group_by(file_name) %>%
  group_split()

batch_size <- 10
track_indices <- seq_along(track_list)
batch_ids <- split(track_indices, ceiling(seq_along(track_indices) / batch_size))

all_batch_files <- character()

for (b in seq_along(batch_ids)) {
  
  cat("Starting batch", b, "of", length(batch_ids), "\n")
  
  this_batch_idx <- batch_ids[[b]]
  this_batch_tracks <- track_list[this_batch_idx]
  batch_results <- vector("list", length(this_batch_tracks))
  
  for (i in seq_along(this_batch_tracks)) {
    
    this_track <- this_batch_tracks[[i]]
    this_name <- this_track$file_name[1]
    safe_name <- make.names(this_name)
    track_outfile <- file.path("crw_outputs/track_files", paste0(safe_name, "_crw.csv"))
    
    if (file.exists(track_outfile)) {
      cat("  Skipping already saved track:", this_name, "\n")
      batch_results[[i]] <- read_csv(track_outfile, show_col_types = FALSE)
      next
    }
    
    cat("  Fitting track", i, "in batch", b, ":", this_name, "\n")
    
    fit_result <- tryCatch(
      fit_one_crw_track(this_track, mod_crw),
      error = function(e) {
        message("    Failed on track: ", this_name)
        message("    Error: ", e$message)
        return(NULL)
      }
    )
    
    batch_results[[i]] <- fit_result
    
    if (!is.null(fit_result)) {
      write_csv(fit_result, track_outfile)
    }
  }
  
  batch_df <- bind_rows(batch_results)
  
  batch_file <- file.path(
    "crw_outputs/batch_files",
    paste0("crw_batch_", sprintf("%03d", b), ".csv")
  )
  
  write_csv(batch_df, batch_file)
  all_batch_files <- c(all_batch_files, batch_file)
  
  cat("Finished batch", b, "- saved to:", batch_file, "\n")
}
```

```{r}
batch_files <- list.files(
  "crw_outputs/batch_files",
  pattern = "^crw_batch_.*\\.csv$",
  full.names = TRUE
)

gps_crw_all <- batch_files %>%
  map_dfr(read_csv, show_col_types = FALSE)

# --------------------------------------------------
# Rejoin trusted metadata from srt_metadata_with_tide_moon
# --------------------------------------------------
meta_tide_moon <- readxl::read_excel(
  "srt_metadata_with_tide_moon.xlsx",
  sheet = "srt_metadata"
) %>%
  transmute(
    file_name = as.character(file_name),
    datetime_srt = as.POSIXct(datetime_srt, tz = "UTC"),
    datetime_local = as.POSIXct(datetime_local, tz = "America/Noronha"),
    tide_height_m,
    tide_state,
    moon_phase,
    moon_illumination
  ) %>%
  distinct(file_name, .keep_all = TRUE)

gps_crw_all <- gps_crw_all %>%
  dplyr::mutate(file_name = as.character(file_name)) %>%
  dplyr::select(-dplyr::any_of(c(
    "datetime_srt", "datetime_local",
    "year", "year.x", "year.y",
    "month_num", "month_num.x", "month_num.y",
    "month", "month.x", "month.y",
    "file_date", "file_date.x", "file_date.y"
  ))) %>%
  dplyr::left_join(meta_tide_moon, by = "file_name") %>%
  dplyr::mutate(
    year = lubridate::year(datetime_local),
    month_num = lubridate::month(datetime_local),
    month = factor(
      lubridate::month(datetime_local, label = TRUE, abbr = TRUE),
      levels = month.abb,
      ordered = TRUE
    ),
    correction_dist_m = sqrt((x_true_m - x_obs)^2 + (y_true_m - y_obs)^2)
  )

# Save the enriched canonical dataset
write_csv(gps_crw_all, "crw_outputs/gps_2s_clean_crw_all_tracks_enriched.csv")


# =========================================
# LOAD CLEAN DATASET
# =========================================
gps_crw_all <- readr::read_csv(
  "crw_outputs/gps_crw_all_enriched_FINAL.csv",
  show_col_types = FALSE
)

# =========================================
#Track-level summary
track_summary <- gps_crw_all %>%
  group_by(file_name) %>%
  summarise(
    n = n(),
    year = first(year),
    month = first(month),
    tide_height_m = first(tide_height_m),
    tide_state = first(tide_state),
    moon_phase = first(moon_phase),
    moon_illumination = first(moon_illumination),
    mean_correction = mean(correction_dist_m, na.rm = TRUE),
    median_correction = median(correction_dist_m, na.rm = TRUE),
    max_correction = max(correction_dist_m, na.rm = TRUE),
    .groups = "drop"
  )

write_csv(track_summary, "crw_outputs/crw_track_summary.csv")

gps_crw_all %>%
  summarise(
    mean_correction = mean(correction_dist_m, na.rm = TRUE),
    median_correction = median(correction_dist_m, na.rm = TRUE),
    max_correction = max(correction_dist_m, na.rm = TRUE)
  )

write_csv(gps_crw_all, "crw_outputs/gps_crw_all_enriched_FINAL.csv")

##data integrity check

gps_crw_all %>%
  dplyr::count(year, month) %>%
  tidyr::pivot_wider(names_from = month, values_from = n, values_fill = 0)
```

## Combine and summarise

```{r}
nrow(gps_crw_all)
n_distinct(gps_crw_all$file_name)

head(track_summary)

summary(track_summary$phi)
summary(track_summary$sigma_gps)
summary(track_summary$sigma_move)
summary(track_summary$mean_correction)
summary(track_summary$median_correction)
summary(track_summary$max_correction)
```

## Corrected hotspot map

```{r}
ggplot(gps_crw_all, aes(x = x_true_m, y = y_true_m)) +
  stat_density_2d_filled(contour_var = "density") +
  coord_equal() +
  theme_bw() +
  labs(
    x = "X (m)",
    y = "Y (m)",
    title = "Tiger shark hotspot areas (Bayesian-corrected positions)"
  )
```

## Cleaned observed hotspot map

```{r}
ggplot(gps_2s_clean, aes(x = longitude, y = latitude)) +
  stat_density_2d_filled(contour_var = "density") +
  coord_equal() +
  theme_bw() +
  labs(
    x = "Longitude",
    y = "Latitude",
    title = "Tiger shark hotspot areas (observed cleaned positions)"
  )
```

## Bayesian-corrected hotspot

```{r}
library(sf)

crs_proj <- 32725

gps_crw_true_sf <- gps_crw_all %>%
  st_as_sf(coords = c("x_true_m", "y_true_m"), crs = crs_proj) %>%   # use your projected CRS object/code here
  st_transform(4326) %>%
  mutate(
    longitude_true = st_coordinates(.)[, 1],
    latitude_true = st_coordinates(.)[, 2]
  )

ggplot() +
  geom_spatraster_rgb(data = sat_map) +
  stat_density_2d_filled(
    data = gps_crw_true_sf,
    aes(x = longitude_true, y = latitude_true),
    contour_var = "density",
    alpha = 0.30,
    bins = 6
  ) +
  coord_sf(
    xlim = st_bbox(bbox_sf)[c("xmin", "xmax")],
    ylim = st_bbox(bbox_sf)[c("ymin", "ymax")],
    expand = FALSE
  ) +
  theme_void() +
  labs(
    title = "Tiger shark core-use areas (Bayesian-corrected positions)",
    fill = "Relative use",
    caption = "Basemap: Esri World Imagery"
  )
```

## Diagnostic comparison figure

```{r}
ggplot() +
  geom_path(
    data = gps_2s_clean,
    aes(x = longitude, y = latitude, group = file_name),
    color = "grey60",
    alpha = 0.20,
    linewidth = 0.15
  ) +
  geom_path(
    data = gps_crw_true_sf,
    aes(x = longitude_true, y = latitude_true, group = file_name),
    color = "red",
    alpha = 0.20,
    linewidth = 0.15
  ) +
  coord_equal() +
  theme_bw() +
  labs(
    x = "Longitude",
    y = "Latitude",
    title = "Observed cleaned paths (grey) and Bayesian-corrected paths (red)"
  )
```

## Year effect and effort

```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(sf)
library(lubridate)

table(gps_crw_all$year)
```

### Effort summary by year

```{r}
effort_year <- gps_crw_all %>%
  group_by(year) %>%
  summarise(
    n_files = n_distinct(file_name),
    n_points = n(),
    mean_track_points = n() / n_distinct(file_name),
    .groups = "drop"
  )

effort_year

ggplot(effort_year, aes(x = year, y = n_files)) +
  geom_col() +
  theme_bw() +
  labs(
    x = "Year",
    y = "Number of SRT files",
    title = "Sampling effort by year"
  )

ggplot(effort_year, aes(x = year, y = n_points)) +
  geom_col() +
  theme_bw() +
  labs(
    x = "Year",
    y = "Number of corrected positions",
    title = "Corrected shark positions by year"
  )
```

### Estimate center of use in meters by year

```{r}
library(sf)

names(gps_crw_all)

gps_crw_all <- gps_crw_all %>%
  st_as_sf(coords = c("x_true_m", "y_true_m"), crs = 32725, remove = FALSE) %>%
  st_transform(4326) %>%
  mutate(
    longitude_true = st_coordinates(.)[, 1],
    latitude_true  = st_coordinates(.)[, 2]
  ) %>%
  st_drop_geometry()

# yearly centers in projected meters
center_year_m <- gps_crw_all %>%
  group_by(year) %>%
  summarise(
    center_x = mean(x_true_m, na.rm = TRUE),
    center_y = mean(y_true_m, na.rm = TRUE),
    n_files = n_distinct(file_name),
    n_points = n(),
    .groups = "drop"
  )

center_year_m

```

### Compare centroids to check if center of use differ by year

```{r}
track_centers <- gps_crw_all %>%
  group_by(file_name, year) %>%
  summarise(
    center_lon = mean(longitude_true, na.rm = TRUE),
    center_lat = mean(latitude_true, na.rm = TRUE),
    center_x = mean(x_true_m, na.rm = TRUE),
    center_y = mean(y_true_m, na.rm = TRUE),
    n_points = n(),
    .groups = "drop"
  )

## test year effect
summary(lm(center_x ~ year, data = track_centers))
summary(lm(center_y ~ year, data = track_centers))
```

```{r}
library(viridis)

year_centroids <- track_centers %>%
  group_by(year) %>%
  summarise(
    mean_x = mean(center_x, na.rm = TRUE),
    mean_y = mean(center_y, na.rm = TRUE),
    .groups = "drop"
  )

track_centers <- track_centers %>%
  mutate(year = factor(year))

year_centroids <- year_centroids %>%
  mutate(year = factor(year))

ggplot(track_centers, aes(x = center_x, y = center_y, color = year, shape = year)) +
  geom_point(size = 2.6, alpha = 0.75) +
  geom_point(
    data = year_centroids,
    aes(x = mean_x, y = mean_y, color = year),
    size = 6,
    shape = 4,
    stroke = 2,
    inherit.aes = FALSE
  ) +
  coord_equal() +
  theme_bw() +
  scale_color_viridis_d(option = "H", end = 0.9) +
  scale_shape_manual(values = c(
    "2022" = 16,
    "2023" = 17,
    "2024" = 15,
    "2025" = 18,
    "2026" = 8
  )) +
  labs(
    x = "X (meters)",
    y = "Y (meters)",
    title = "Track-level centers of use by year",
    color = "Year",
    shape = "Year"
  )
```

### Effort by year bar plot

```{r}
ggplot(effort_year, aes(x = year, y = n_files)) +
  geom_col() +
  theme_bw() +
  labs(
    x = "Year",
    y = "Number of SRT files",
    title = "Sampling effort by year"
  )
```

## Overall hotspot figure - density in projected space (meters)

```{r}
library(dplyr)
library(sf)
library(terra)
library(MASS)
library(ggplot2)
library(tidyterra)

turbo_dark <- c(
  "#30123B",
  "#2C7FB8",
  "#4AC16D",
  "#F2C14E",
  "#C81D11"
)

# make sure projected corrected coordinates exist
# use gps_crw_all for corrected positions
x_range <- range(gps_crw_all$x_true_m, na.rm = TRUE)
y_range <- range(gps_crw_all$y_true_m, na.rm = TRUE)

kde_m <- MASS::kde2d(
  x = gps_crw_all$x_true_m,
  y = gps_crw_all$y_true_m,
  n = 300,
  h = c(25, 25),
  lims = c(x_range, y_range)
)

kde_m_df <- expand.grid(
  x = kde_m$x,
  y = kde_m$y
)
kde_m_df$z <- as.vector(kde_m$z)

vals_m <- kde_m_df$z
q_m <- quantile(vals_m, probs = seq(0, 1, length.out = 6), na.rm = TRUE)

q_m[1] <- min(vals_m, na.rm = TRUE) - 1e-12
q_m[length(q_m)] <- max(vals_m, na.rm = TRUE) + 1e-12

kde_m_df <- kde_m_df %>%
  mutate(
    use_class = cut(
      z,
      breaks = q_m,
      include.lowest = TRUE,
      labels = c("Very low", "Low", "Moderate", "High", "Very high")
    )
  )

p_hotspot_overall <- ggplot(kde_m_df, aes(x = x, y = y, fill = use_class)) +
  geom_raster() +
  coord_equal() +
  scale_fill_manual(
    values = turbo_dark,
    drop = FALSE,
    name = "Relative use"
  ) +
  theme_bw(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 10)
  ) +
  labs(
    x = "X (m)",
    y = "Y (m)",
    title = "Tiger shark core-use areas"
  )

p_hotspot_overall
```

## Hotspot by year

## Kernel density estimation overlayed on basemap hotspot figure

```{r}

turbo_dark <- c(
  "#30123B",
  "#2C7FB8",
  "#4AC16D",
  "#F2C14E",
  "#C81D11"
)

# --------------------------------------------------
# 1. KDE in projected space (meters)
# --------------------------------------------------
x_range_m <- range(gps_crw_all$x_true_m, na.rm = TRUE)
y_range_m <- range(gps_crw_all$y_true_m, na.rm = TRUE)

kde_m <- MASS::kde2d(
  x = gps_crw_all$x_true_m,
  y = gps_crw_all$y_true_m,
  n = 300,
  h = c(25, 25),
  lims = c(x_range_m, y_range_m)
)

# --------------------------------------------------
# 2. Convert KDE grid to raster in UTM 25S
# --------------------------------------------------
kde_m_df <- expand.grid(
  x = kde_m$x,
  y = kde_m$y
)
kde_m_df$z <- as.vector(kde_m$z)

kde_rast_utm <- rast(
  kde_m_df,
  type = "xyz",
  crs = "EPSG:32725"
)

names(kde_rast_utm) <- "density"

# --------------------------------------------------
# 3. Reproject to lon/lat
# --------------------------------------------------
kde_rast_ll <- terra::project(kde_rast_utm, "EPSG:4326")
kde_rast_ll[kde_rast_ll < quantile(values(kde_rast_ll), 0.05, na.rm = TRUE)] <- NA

# --------------------------------------------------
# 4. Convert to data frame for plotting
# --------------------------------------------------
# trim only the weakest background
vals <- terra::values(kde_rast_ll)[, 1]
thresh <- quantile(vals[vals > 0], 0.15, na.rm = TRUE)

kde_trim <- kde_rast_ll
kde_trim[kde_trim < thresh] <- NA

# convert to data frame and drop NA cells
kde_df_plot <- as.data.frame(kde_trim, xy = TRUE, na.rm = TRUE)
names(kde_df_plot)[3] <- "density"

# --------------------------------------------------
# 5. Basemap + filled contours
# --------------------------------------------------
p_basemap_context <- ggplot() +
  geom_spatraster_rgb(data = sat_map) +
  geom_raster(
    data = kde_df_plot,
    aes(x = x, y = y, fill = density),
    alpha = 0.65
  ) +
  coord_sf(
    xlim = st_bbox(bbox_sf)[c("xmin", "xmax")],
    ylim = st_bbox(bbox_sf)[c("ymin", "ymax")],
    expand = FALSE
  ) +
  scale_fill_gradientn(
    colours = turbo_dark,
    values = scales::rescale(c(0, 0.05, 0.15, 0.40, 1)),
    name = "Relative use",
    breaks = c(
    min(kde_df_plot$density, na.rm = TRUE),
    max(kde_df_plot$density, na.rm = TRUE)
  ),
  labels = c("Low", "Very high")
) +
guides(
  fill = guide_colorbar(reverse = TRUE)
    ) +
  theme_void(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.title = element_text(face = "bold"),
    legend.background = element_rect(fill = "white", colour = "grey80")
  ) +
  labs(
    title = "Tiger shark core-use areas",
    caption = "Basemap: Esri World Imagery"
  )

p_basemap_context
```

### Test for month effect

```{r}
gps_crw_all %>%
  dplyr::count(year, month) %>%
  tidyr::pivot_wider(names_from = month, values_from = n, values_fill = 0)

gps_month <- gps_crw_all %>%
  dplyr::filter(!is.na(year), !is.na(month))
```

```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(scales)
library(grid)

turbo_dark <- c(
  "#30123B",
  "#2C7FB8",
  "#4AC16D",
  "#F2C14E",
  "#C81D11"
)

# -----------------------------------------
# Use cleaned dataset as-is
# -----------------------------------------
gps_month <- gps_crw_all %>%
  dplyr::filter(
    !is.na(month),
    !is.na(month_num),
    !is.na(year)
  )

#check Year X Month sampling balance
gps_month %>%
  dplyr::count(year, month) %>%
  tidyr::pivot_wider(names_from = month, values_from = n, values_fill = 0)

# optional checks
table(gps_month$month, useNA = "ifany")
table(gps_month$year, useNA = "ifany")

effort_month <- gps_month %>%
  group_by(month, month_num) %>%
  summarise(
    n_files = n_distinct(file_name),
    n_points = n(),
    mean_track_points = n() / n_distinct(file_name),
    .groups = "drop"
  ) %>%
  arrange(month_num)

print(effort_month)

p_effort_month_files <- ggplot(effort_month, aes(x = month, y = n_files)) +
  geom_col(fill = "grey40") +
  theme_bw(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold")
  ) +
  labs(
    x = "Month",
    y = "Number of SRT files",
    title = "Sampling effort by month"
  )

p_effort_month_points <- ggplot(effort_month, aes(x = month, y = n_points)) +
  geom_col(fill = "grey40") +
  theme_bw(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold")
  ) +
  labs(
    x = "Month",
    y = "Number of corrected positions",
    title = "Corrected shark positions by month"
  )

print(p_effort_month_files)
print(p_effort_month_points)

p_hotspot_month <- ggplot(gps_month, aes(x = longitude_true, y = latitude_true)) +
  stat_density_2d_filled(
    contour_var = "density",
    bins = 5
  ) +
  coord_equal() +
  facet_wrap(~ month) +
  scale_fill_manual(
    values = turbo_dark,
    name = "Relative use",
    labels = c("Very low", "Low", "Moderate", "High", "Very high")
  ) +
  scale_x_continuous(
    breaks = pretty_breaks(n = 2),
    labels = label_number(accuracy = 0.001)
  ) +
  scale_y_continuous(
    breaks = pretty_breaks(n = 3),
    labels = label_number(accuracy = 0.001)
  ) +
  theme_bw(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    panel.spacing = unit(1, "lines"),
    plot.title = element_text(face = "bold", hjust = 0.5),
    strip.text = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 10)
  ) +
  labs(
    x = "Longitude",
    y = "Latitude",
    title = "Tiger shark hotspot areas by month"
  )

print(p_hotspot_month)

track_centers_month <- gps_month %>%
  group_by(file_name, month, month_num) %>%
  summarise(
    center_lon = mean(longitude_true, na.rm = TRUE),
    center_lat = mean(latitude_true, na.rm = TRUE),
    center_x = mean(x_true_m, na.rm = TRUE),
    center_y = mean(y_true_m, na.rm = TRUE),
    n_points = n(),
    .groups = "drop"
  ) %>%
  arrange(month_num)

print(head(track_centers_month))

month_centroids <- track_centers_month %>%
  group_by(month, month_num) %>%
  summarise(
    mean_x = mean(center_x, na.rm = TRUE),
    mean_y = mean(center_y, na.rm = TRUE),
    n_files = n(),
    .groups = "drop"
  ) %>%
  arrange(month_num)

p_centers_month <- ggplot(track_centers_month, aes(x = center_x, y = center_y, color = month, shape = month)) +
  geom_point(size = 2.4, alpha = 0.75) +
  geom_point(
    data = month_centroids,
    aes(x = mean_x, y = mean_y, color = month),
    shape = 4,
    stroke = 2,
    size = 5.5,
    inherit.aes = FALSE
  ) +
  coord_equal() +
  scale_color_viridis_d(option = "C", end = 0.9, name = "Month") +
  theme_bw(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 10)
  ) +
  labs(
    x = "X (meters)",
    y = "Y (meters)",
    title = "Track-level centers of use by month"
  )

print(p_centers_month)

p_center_y_month <- ggplot(track_centers_month, aes(x = month, y = center_y, fill = month)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_viridis_d(option = "C", end = 0.9, guide = "none") +
  theme_bw(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold")
  ) +
  labs(
    x = "Month",
    y = "Track center Y (meters)",
    title = "North–south center of shark space use by month"
  )

p_center_x_month <- ggplot(track_centers_month, aes(x = month, y = center_x, fill = month)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_viridis_d(option = "C", end = 0.9, guide = "none") +
  theme_bw(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold")
  ) +
  labs(
    x = "Month",
    y = "Track center X (meters)",
    title = "East–west center of shark space use by month"
  )

print(p_center_y_month)
print(p_center_x_month)

lm_month_y <- lm(center_y ~ month, data = track_centers_month)
lm_month_x <- lm(center_x ~ month, data = track_centers_month)

summary(lm_month_y)
summary(lm_month_x)

month_summary <- track_centers_month %>%
  group_by(month, month_num) %>%
  summarise(
    n_files = n(),
    mean_center_x = mean(center_x, na.rm = TRUE),
    mean_center_y = mean(center_y, na.rm = TRUE),
    sd_center_x = sd(center_x, na.rm = TRUE),
    sd_center_y = sd(center_y, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(month_num)

print(month_summary)

write_csv(effort_month, "effort_by_month.csv")
write_csv(track_centers_month, "track_centers_by_month.csv")
write_csv(month_summary, "month_center_summary.csv")
```

### GAM

```{r}
library(mgcv)

track_centers_month <- track_centers_month %>%
  dplyr::left_join(
    gps_crw_all %>%
      dplyr::select(file_name, year) %>%
      distinct(),
    by = "file_name"
  ) %>%
  dplyr::mutate(
    month = factor(month, levels = month.abb, ordered = TRUE),
    year = factor(year)
  )

#check data structure
str(track_centers_month)

#GAM model North-South
gam_month_y <- mgcv::gam(
  center_y ~ month + s(year, bs = "re"),
  data = track_centers_month,
  method = "REML"
)

summary(gam_month_y)

#GAM model East-West
gam_month_x <- mgcv::gam(
  center_x ~ month + s(year, bs = "re"),
  data = track_centers_month,
  method = "REML"
)

summary(gam_month_x)
```

## GAM Temporal and environmental drivers

```{r}
#check data
names(gps_crw_all)

# 1. Build track-level environmental data
# --------------------------------------------------
env_track <- gps_crw_all %>%
  dplyr::group_by(file_name) %>%
  dplyr::summarise(
    tide_height_m = mean(tide_height_m, na.rm = TRUE),
    tide_state = names(sort(table(tide_state), decreasing = TRUE))[1],
    moon_phase = names(sort(table(moon_phase), decreasing = TRUE))[1],
    moon_illumination = mean(moon_illumination, na.rm = TRUE),
    .groups = "drop"
  )

# --------------------------------------------------
# 2. Rebuild track-level analysis dataset from scratch
#    IMPORTANT: month must be an unordered factor here
# --------------------------------------------------
track_centers_env <- track_centers_month %>%
  dplyr::select(
    file_name, month, month_num, center_lon, center_lat,
    center_x, center_y, n_points, year
  ) %>%
  dplyr::left_join(env_track, by = "file_name") %>%
  dplyr::mutate(
    month = factor(as.character(month), levels = month.abb, ordered = FALSE),
    year = factor(as.character(year)),
    tide_state = factor(tide_state),
    moon_phase = factor(moon_phase)
  ) %>%
  droplevels()

# --------------------------------------------------
# 3. Keep only complete cases needed for this model
# --------------------------------------------------
track_centers_env <- track_centers_env %>%
  dplyr::filter(
    !is.na(center_y),
    !is.na(center_x),
    !is.na(month),
    !is.na(year),
    !is.na(tide_height_m),
    !is.na(moon_illumination)
  ) %>%
  droplevels()

# --------------------------------------------------
# 4. Check actual factor levels used in the model
# --------------------------------------------------
print(levels(track_centers_env$month))
print(levels(track_centers_env$year))
```

```{r}
#| cache: true

##North-south (center_y) GAM
gam_y_full <- mgcv::gam(
  center_y ~ month + s(year, bs = "re") + s(tide_height_m) + s(moon_illumination),
  data = track_centers_env,
  method = "REML"
)

## East-West (center_x) GAM
gam_x_full <- mgcv::gam(
  center_x ~ month + s(year, bs = "re") + s(tide_height_m) + s(moon_illumination),
  data = track_centers_env,
  method = "REML"
)
```

```{r}
#| cache: true

summary(gam_y_full)
summary(gam_x_full)

#GAM diagnostics
gam.check(gam_y_full)
gam.check(gam_x_full)

#Residuals diagnostics
par(mfrow = c(2,2))
plot(gam_y_full, residuals = TRUE, pch = 16)
plot(gam_x_full, residuals = TRUE, pch = 16)
par(mfrow = c(1,1))
```

```{r}
#| cache: false
#| include: false

## Build prediction using year 2024 - year with the broader sampling coverage
ref_year <- "2024"
if (!ref_year %in% levels(track_centers_env$year)) {
  ref_year <- levels(track_centers_env$year)[1]
}

pred_month <- data.frame(
  month = factor(levels(track_centers_env$month), levels = levels(track_centers_env$month)),
  year = factor(ref_year, levels = levels(track_centers_env$year)),
  tide_height_m = mean(track_centers_env$tide_height_m, na.rm = TRUE),
  moon_illumination = mean(track_centers_env$moon_illumination, na.rm = TRUE)
)

cat("pred_month built successfully\n")
```

## 2-seconds GPS locations GAM with random effect

```{r}
##create clean 2-sec dataset before modelling
gps_env <- gps_crw_all %>%
  dplyr::filter(
    !is.na(y_true_m),
    !is.na(x_true_m),
    !is.na(month),
    !is.na(year),
    !is.na(tide_height_m),
    !is.na(moon_illumination),
    !is.na(file_name)
  ) %>%
  dplyr::mutate(
    month = factor(as.character(month), levels = month.abb),
    year = factor(year),
    file_name = factor(file_name)
  ) %>%
  droplevels()

##sanity check
dplyr::glimpse(gps_env)
```

```{r}
## 2 seconds GAM model
gam_2s_y <- mgcv::bam(
  y_true_m ~ 
    s(x_true_m, k = 20) +
    month +
    s(tide_height_m, k = 10) +
    s(moon_illumination, k = 10) +
    s(year, bs = "re") +
    s(file_name, bs = "re"),
  data = gps_env,
  method = "fREML",
  discrete = TRUE
)

saveRDS(gam_2s_y, "gam_2s_y.rds")
```

## Relevant statistics

```{r}
cat("GAM Y:\n")
cat("Deviance explained:", summary(gam_y_full)$dev.expl, "\n")
cat("Adjusted R2:", summary(gam_y_full)$r.sq, "\n\n")

cat("GAM X:\n")
cat("Deviance explained:", summary(gam_x_full)$dev.expl, "\n")
cat("Adjusted R2:", summary(gam_x_full)$r.sq, "\n")
```

## Figures set

### Figure 1 — Spatial context

```{r}
p_basemap_context <- ggplot() +
  geom_spatraster_rgb(data = sat_map) +
  geom_raster(
    data = kde_df_plot,
    aes(x = x, y = y, fill = density),
    alpha = 0.65
  ) +
  coord_sf(
    xlim = st_bbox(bbox_sf)[c("xmin", "xmax")],
    ylim = st_bbox(bbox_sf)[c("ymin", "ymax")],
    expand = FALSE
  ) +
  scale_fill_gradientn(
  colours = turbo_dark,
  values = scales::rescale(c(0, 0.05, 0.15, 0.40, 1)),
  name = "Relative space use",
  breaks = c(
    min(kde_df_plot$density, na.rm = TRUE),
    max(kde_df_plot$density, na.rm = TRUE)
  ),
  labels = c("Low", "High"),
  guide = guide_colorbar(
    reverse = TRUE,
    barheight = unit(3.5, "cm"),
    barwidth  = unit(0.4, "cm")
      )
  ) +
  theme_void(base_size = 13) +
  theme(
    legend.position = "right",
    legend.key.height = unit(0.5, "cm"),
    legend.key.width = unit(0.4, "cm"),
    legend.title = element_text(size = 10.5, face = "bold"),
    legend.text = element_text(size = 9.5),
    legend.background = element_rect(fill = "white", colour = "grey80")
  ) +
  labs(
    title = "Tiger shark core-use areas",
    caption = "Basemap: Esri World Imagery"
  )

p_basemap_context
```

### Figure 2: Predicted month effect

```{r}
pred_month <- data.frame(
  month = factor(levels(track_centers_env$month), levels = levels(track_centers_env$month)),
  month_num = match(levels(track_centers_env$month), month.abb),  # 1–12
  year = factor(ref_year, levels = levels(track_centers_env$year)),
  tide_height_m = mean(track_centers_env$tide_height_m, na.rm = TRUE),
  moon_illumination = mean(track_centers_env$moon_illumination, na.rm = TRUE)
)

#North–south (center_y)
pred_month_y$month_num <- match(pred_month_y$month, month.abb)

p_month_y <- ggplot(pred_month_y, aes(x = month_num, y = fit, group = 1)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_bw(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold")
  ) +
  scale_x_continuous(breaks = 1:12) +
  labs(
    x = "Month",
    y = "North–south position (UTM, m)",
    title = "North–south shift"
  )

#East-west (center_x)
pred_month_x$month_num <- match(pred_month_x$month, month.abb)

p_month_x <- ggplot(pred_month_x, aes(x = month_num, y = fit, group = 1)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_bw(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold")
  ) +
  scale_x_continuous(breaks = 1:12) +
  labs(
    x = "Month",
    y = "East–west position (UTM, m)",
    title = "East–west shift"
  )


#combine p_month_y and p_month_x
library(patchwork)

figure_month <- p_month_y + p_month_x +
  plot_layout(ncol = 2) +
  plot_annotation(
    title = "Seasonal variation in shark space use",
    tag_levels = "A",
    theme = theme(
      plot.title = element_text(face = "bold", hjust = 0.5)
    )
  )

figure_month

ggsave(
  filename = "figure_month_effects.png",
  plot = figure_month,
  width = 11,
  height = 5.5,
  dpi = 300
)
```

```{r}
## Predict center_y with confidence intervals
pred_y <- predict(gam_y_full, newdata = pred_month, type = "link", se.fit = TRUE)

pred_month_y <- pred_month %>%
  dplyr::mutate(
    fit = pred_y$fit,
    se = pred_y$se.fit,
    lower = fit - 1.96 * se,
    upper = fit + 1.96 * se,
    month_num = match(month, month.abb)
  )


## Plot predicted month effect for center_y
p_month_y <- ggplot(pred_month_y, aes(x = month, y = fit, group = 1)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_bw(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold")
  ) +
  labs(
    x = "Month",
    y = "Predicted center_y",
    title = "Predicted north–south shark space use by month"
  )

print(p_month_y)
```

```{r}
## Predict center_x with confidence intervals
pred_x <- predict(gam_x_full, newdata = pred_month, type = "link", se.fit = TRUE)

pred_month_x <- pred_month %>%
  dplyr::mutate(
    fit = pred_y$fit,
    se = pred_y$se.fit,
    lower = fit - 1.96 * se,
    upper = fit + 1.96 * se,
    month_num = match(month, month.abb)
  )


## Plot predicted month effect for center_x
p_month_x <- ggplot(pred_month_x, aes(x = month, y = fit, group = 1)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_bw(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold")
  ) +
  labs(
    x = "Month",
    y = "Predicted center_x",
    title = "Predicted east–west shark space use by month"
  )

print(p_month_x)
```

### Figure 3: GAM smooths (tide+moon+year)

```{r}
library(mgcv)

# save current graphics settings so you can restore them after
old_par <- par(no.readonly = TRUE)

# 2 x 2 layout
par(
  mfrow = c(2, 2),
  mar = c(4.5, 4.5, 3, 1),
  oma = c(0, 0, 2, 0),
  cex.main = 1.1,
  cex.lab = 1.1,
  cex.axis = 1
)

# A. Tide effect on north–south position
plot(
  gam_y_full,
  select = 2,   # s(tide_height_m)
  shade = TRUE,
  shade.col = "grey80",
  seWithMean = TRUE,
  rug = TRUE,
  residuals = FALSE,
  main = "A. Tide effect on north–south position",
  xlab = "Tide height (m)",
  ylab = "Effect on position (m)"
)

# B. Moon effect on north–south position
plot(
  gam_y_full,
  select = 3,   # s(moon_illumination)
  shade = TRUE,
  shade.col = "grey80",
  seWithMean = TRUE,
  rug = TRUE,
  residuals = FALSE,
  main = "B. Moon effect on north–south position",
  xlab = "Moon illumination",
  ylab = "Effect on position (m)"
)

# C. Tide effect on east–west position
plot(
  gam_x_full,
  select = 2,   # s(tide_height_m)
  shade = TRUE,
  shade.col = "grey80",
  seWithMean = TRUE,
  rug = TRUE,
  residuals = FALSE,
  main = "C. Tide effect on east–west position",
  xlab = "Tide height (m)",
  ylab = "Effect on position (m)"
)

# D. Moon effect on east–west position
plot(
  gam_x_full,
  select = 3,   # s(moon_illumination)
  shade = TRUE,
  shade.col = "grey80",
  seWithMean = TRUE,
  rug = TRUE,
  residuals = FALSE,
  main = "D. Moon effect on east–west position",
  xlab = "Moon illumination",
  ylab = "Effect on position (m)"
)

# overall title
mtext(
  "Environmental effects on tiger shark space use",
  outer = TRUE,
  cex = 1.4,
  font = 2
)

# restore previous graphics settings
par(old_par)

png("figure_3_environmental_effects.png", width = 2000, height = 1600, res = 300)

old_par <- par(no.readonly = TRUE)

par(
  mfrow = c(2, 2),
  mar = c(4.5, 4.5, 3, 1),
  oma = c(0, 0, 2, 0),
  cex.main = 1.1,
  cex.lab = 1.1,
  cex.axis = 1
)

plot(gam_y_full, select = 2, shade = TRUE, shade.col = "grey80",
     seWithMean = TRUE, rug = TRUE, residuals = FALSE,
     main = "A. Tide effect on north–south position",
     xlab = "Tide height (m)", ylab = "Effect on position (m)")

plot(gam_y_full, select = 3, shade = TRUE, shade.col = "grey80",
     seWithMean = TRUE, rug = TRUE, residuals = FALSE,
     main = "B. Moon effect on north–south position",
     xlab = "Moon illumination", ylab = "Effect on position (m)")

plot(gam_x_full, select = 2, shade = TRUE, shade.col = "grey80",
     seWithMean = TRUE, rug = TRUE, residuals = FALSE,
     main = "C. Tide effect on east–west position",
     xlab = "Tide height (m)", ylab = "Effect on position (m)")

plot(gam_x_full, select = 3, shade = TRUE, shade.col = "grey80",
     seWithMean = TRUE, rug = TRUE, residuals = FALSE,
     main = "D. Moon effect on east–west position",
     xlab = "Moon illumination", ylab = "Effect on position (m)")

mtext("Environmental effects on shark space use", outer = TRUE, cex = 1.4, font = 2)

par(old_par)
dev.off()
```
